---
title: "Homabay - HIV/AIDS Modelling"
author: "Bonsoul"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r,libararies}
library(ggplot2)
library(tidyverse)
library(plotly)
library(rlang)
library(psych)
library(knitr)
library(stringr)
library(questionr)
library(gtsummary)
library(shiny)
library(lubridate)
library(magrittr)
library(expss)
library(rnaturalearth)
library(rnaturalearthdata)
library(CARBayes)
library(gridExtra)             
library(grid) 
library(survival)
library(caret) 
library(randomForest) 
library(dplyr) 
library(corrplot) 
library(readxl) 
library(tidyr) 
library(survminer) 
library(sf) 
library(tmap)
library(RColorBrewer)
library(kableExtra)
library(broom)
library(spdep)
library(brglm2)
```

```{r,reading data}
df = read_excel("D:/Downloads/HIV - VARIABLES.xlsx")
View(df)
```


```{r,EDA}
#Rename coordinate columns for clarity
df1 <- df %>% rename(latitude = var52, longitude = var51)

duplicates <- df1[duplicated(df1),] #checking duplicates
print(duplicates)


missing_per_column <- colSums(is.na(df1)) #columns with null values
total_missing <- sum(missing_per_column) 
print(total_missing) 
print(missing_per_column)

columns_with_missing <- names(missing_per_column[missing_per_column > 0]) 
print(columns_with_missing)
```
```{r Summary statistics for numerical variables }
#Compute summary statistics for numeric columns
summary_stats <- df1 %>% 
  select_if(is.numeric) %>% 
  summarise(across(
    everything(), 
    list(
      min = ~min(., na.rm = TRUE),
      Q1 = ~quantile(., 0.25, na.rm = TRUE),
      median = ~median(., na.rm = TRUE),
      mean = ~mean(., na.rm = TRUE),
      Q3 = ~quantile(., 0.75, na.rm = TRUE),
      max = ~max(., na.rm = TRUE),
      sd = ~sd(., na.rm = TRUE)
    )
  ))
summary(summary_stats)
```

```{r  Frequency distribution for categorical variables}
categorical_vars <- df1 %>%
  select_if(is.character)

for (var in names(categorical_vars)) {
  print(var)
  print(table(categorical_vars[[var]]))
}
```

```{r,hitograms}
#Visualize distributions with histograms

ggplot(df1, aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) + 
  ggtitle("Histogram of Age") + 
  theme_minimal()
  
  
ggplot(df1, aes(x = weightkg)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) + 
  ggtitle("Histogram of weightkg") + 
  theme_minimal()
  

ggplot(df1, aes(x = durationofbfmonths)) + 
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) + 
  ggtitle("Histogram of durationofbfmonths") + 
  theme_minimal()
```


```{r,boxplots}
# Define numeric variables for boxplots
numeric_vars <- c("age", "weightkg", "bwtkgs", "babywt6wks", "time", "_t")

# Create boxplots for each numeric variable
for (var in numeric_vars) {
  p <- ggplot(df1, aes(x = monthsantibodytest, y = .data[[var]], fill = monthsantibodytest)) +
    geom_boxplot() +
    labs(title = paste("Boxplot of", var, "by Months Antibody Test"),
         x = "Months Antibody Test",
         y = var) +
    theme_minimal()
  
  print(p)
}
```

```{r,gender distribution}
# Create a bar chart for gender distribution
gender_plot <- ggplot(df1, aes(x = sexofthebaby, fill = sexofthebaby)) +
  geom_bar() +
  labs(title = "Gender Distribution",
       x = "Gender",
       y = "Count") +
  theme_minimal()

print(gender_plot)
```
```{r,density plots}
# Create density plots
for (var in numeric_vars) {
  density_plot <- ggplot(df1, aes(x = .data[[var]], color = monthsantibodytest)) +
    geom_density() +
    labs(title = paste("Density Plot of", var), x = var, y = "Density") +
    theme_minimal()
  
  print(density_plot)
}
```

```{r,Mutating +ve and -ne to be numerical}
df2 <- df1 %>% mutate(monthsantibodytest = ifelse(tolower(monthsantibodytest) == "positive", 1, 0))

table(df1$monthsantibodytest)
```

### socio demo characteristic of parents

```{r franc1,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(subcounty, maritalstatus, residence, employmentstatus, educationlevel, monthsantibodytest) %>%
  tbl_summary(by = monthsantibodytest) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### ANC attendance of mothers


### ANC attendance of mothers
```{r franc2,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(ancattendance, facilitylevelanc, distancetofacility, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    type = everything() ~ "categorical"
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```



### gestation age at first anc visits wks

```{r franc3,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(gestationatfirstancwks, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    type = gestationatfirstancwks ~ "continuous2",
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### hiv status before pregnancy and syphillis

```{r franc4,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(hivstatusbeforepregnancy, syphillis, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### uti, miscarriage

```{r franc5,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(uti, miscarriage, abortion, stillbirth, premature, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### HIV and CD4

```{r franc6,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(haartduringpregnancy, cd4cellcountcellsmm3, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### mother history of having sexually transmitted disease during pregnancy

```{r franc7,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(historyofstiduringpregnancy, genitalwart, vaginaldischargesyndrome, genitalherpes, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### treatedduringpregnancy

```{r franc8,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(treatedduringpregnancy, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### malaria, anaemia, hypertention

```{r franc9,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(malaria, anaemia, hypertention, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

### numberofchildren

```{r franc10,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(numberofchildren, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    type = numberofchildren ~ "categorical",
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### patnershivstatus, whohivdiseasestage

```{r franc12,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(patnershivstatus, whohivdiseasestage, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()

```

### numberofancvisitsmade

```{r franc13,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr::select(numberofancvisitsmade, monthsantibodytest) %>%
  tbl_summary(
    by=monthsantibodytest,
    type = numberofancvisitsmade ~ "categorical",
    statistic = list(all_categorical() ~ "{p}% ({n}/{N})",
                  all_dichotomous() ~ "{p}% ({n}/{N})",
                     all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})"))
   ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

### hemoglobinlevelduringpregnancy

```{r franc14,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr:: select(hemoglobinlevelduringpregnancy, monthsantibodytest) %>%
  tbl_summary(
    by=monthsantibodytest,
    type = hemoglobinlevelduringpregnancy ~ "continuous2",
    statistic = list(all_categorical() ~ "{p}% ({n}/{N})",
                  all_dichotomous() ~ "{p}% ({n}/{N})",
                     all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})"))
   ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

### durationofbfmonths

```{r franc15,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr:: select(lduration,durationofbfmonths, monthsantibodytest) %>%
  tbl_summary(
    by=monthsantibodytest,
    type = durationofbfmonths ~ "continuous2",
    statistic = list(all_categorical() ~ "{p}% ({n}/{N})",
                  all_dichotomous() ~ "{p}% ({n}/{N})",
                     all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})"))
   ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

### sexofthebaby

```{r franc16,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr:: select(sexofthebaby, monthsantibodytest) %>%
  tbl_summary(
    by=monthsantibodytest,
   # type = durationofbfmonths ~ "continuous2",
    statistic = list(all_categorical() ~ "{p}% ({n}/{N})",
                  all_dichotomous() ~ "{p}% ({n}/{N})",
                     all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})"))
   ) %>%
  add_overall() %>%
  add_p() %>%
  bold_labels()
```

#subcounty

```{r franc17,echo=FALSE, message=FALSE, warning=FALSE, include=T}
df2 %>%
  dplyr:: select(subcounty, monthsantibodytest) %>%
  tbl_summary(
    by = monthsantibodytest,  # Grouping by monthsantibodytest
    statistic = list(
      all_categorical() ~ "{p}% ({n}/{N})",
      all_dichotomous() ~ "{p}% ({n}/{N})",
      all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})")
    )
  ) %>%
  add_overall() %>%  # Now valid because `by` is specified
  add_p() %>%
  bold_labels()

```


### PCR test trend

```{r franc17,echo=FALSE, message=FALSE, warning=FALSE, include=T}
data_fr %>%
  select(birthpcr,weekspcr, monthspcr,yearpcr,monthsantibodytest) %>%
  tbl_summary(
   # by=monthsantibodytest,
   # type = durationofbfmonths ~ "continuous2",
    statistic = list(all_categorical() ~ "{p}% ({n}/{N})",
                  all_dichotomous() ~ "{p}% ({n}/{N})",
                     all_continuous2() ~ c("{median} ({min} - {max})", "{mean} ({sd})"))
   ) %>%
  #add_overall() %>%
 # add_p() %>%
  bold_labels()
```